Data Migration under a Changing Schema in Ampersand

### 2024-05-25 2024_RAMiCS_Migration/articleMigrationRAMiCS.tex
transfer Migration paper to RAMiCS

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Schema changes cannot always be avoided when updating software, so a \define{schema-changing data migration} (SCDM) will be necessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, ETL tools typically provide little support for invariants that change, forcing development teams to write code.
   In a world where automation of the software process is resulting in higher productivity, more frequent releases, and better quality,
   SCDM's should not stay behind.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   The risk and effort of SCDMs explains why these teams try to avoid schema changes in the first place.
   We believe that software teams that realize high update frequencies need better tools to perform SCDM's reliably with zero-downtime.
   We want then to design, implement, and deploy SCDMs in the flow of their DevOps practice.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes
   to avoid introducing new errors in an otherwise error-prone migration process.
   In another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or an implicit schema, e.g.~\cite{Hillenbrand2022}.
   Such situations lack the error preventing power that explicit schemas bring during the development of software.
   All errors a schema can prevent at compile time must then be compensated by runtime checks,
   which increases the likelyhood of end-users getting error messages.
   This requires versioned storage of production data and an overhead in performance.
   That is why this paper focuses on SCDMs for systems with an explicit schema.
   The prototypical use case for that is to release updates of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

@@

   The next section analyzes SCDMs with an eye on zero downtime and data quality.
   It sketches the outline of a procedure for SCDMs.
   Section~\ref{sct:Definitions} formalizes the concepts that we need to define the procedure.
   Section~\ref{sct:Generating} defines the migration system that automates SCDMs.
   Section~\ref{sct:PoC} demonstrates the prototype of a migration system, which we used to verify our approach experimentally.
   We have used the language Ampersand for this purpose,
   because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
   Finally, section~\ref{sct:Validation} discusses the validation of our approach by showing that all requirements are met.

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
   The current section yields a procedure for migrating data from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   \begin{figure}[bht]
      \begin{center}
        \includegraphics[scale=0.8]{figures/existing_system.pdf}
      \end{center}
   \caption{Anatomy of an information system}
   \label{fig:pre-migration}
   \end{figure}

   Figure~\ref{fig:pre-migration} depicts the situation before migration.
   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
   Our research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
   To keep our approach technology independent, we assume that data sets contain triples.
   This makes our approach valid for every kind of database that triples can represent,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.

   We assume that constraints implement the business semantics of the data.
   Constraints represent business concerns formally, so they can be checked automatically and can be used to generate software.
   Some of the constraints require human intervention, while others require a system to intervene.
   In this paper, we distinguish three different kinds of constraints:
\begin{enumerate}
\item Blocking invariant\\
   A \define{blocking invariant} is a constraint that is always satisfied in a system.
   It serves to constrain the data set at runtime.
   When the data set changes in a way that violates a blocking invariant, the system produces an error message and refuses the change.
\item Transactional invariant\\
   The classical database transaction can be understood as a \define{transactional invariant},
   which the system keeps satisfied by adding or deleting triples to the data set.
   As soon as the data violates this constraint, the system restores it without human intervention.
   So, the outside world experiences this as a constraint that is always satisfied, i.e.~an invariant.
\item Business constraint\\
   A \define{business constraint} is a constraint that users can violate temporarily until someone restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").
   That takes some time, during which the constraint is violated.
   So, we do not consider business constraints to be invariants.
\end{enumerate}
   Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
   Of the three types of constraint, only two are invariants.

\subsection{Ampersand}
   We employ Ampersand as a prototyping language to demonstrate our approach.
   Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our approach.
   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
   The systems they generate keep invariants satisfied and alert users to violations of business constraints.
   The absence of imperative code in Ampersand scripts enhances reasoning about the system,
   while its static typing~\cite{vdWoude2011} yields the established benefits in software engineering processes~\cite{HanenbergKRTS14,Petersen2014}.
   Constraints carry business semantics, which makes ``preserving the meaning as much as possible'' explicit.
   An Ampersand script provides just enough information to generate a complete system,
   allowing extraction of a classical database schema (i.e., data structure plus constraints) from the Ampersand script.

@@

   Atoms serve as data elements.
   All atoms are taken from an infinite set called $\Atoms$.
   %
   Concepts are names that group atoms of the same type.
   All concepts are taken from an infinite set $\Concepts$.
   %$\Concepts$ and $\Atoms$ are disjoint.
   For example, a developer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
   and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \concept{Person} and \concept{TelephoneNumber} are concepts.
   Moreover, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms.
   In the sequel, variables $A$, $B$, and $C$ will represent concepts, and variables $a$, and $b$ represent \emph{atoms}.
   %
   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
   The term $a\instance C$ means that atom $a$ is an \emph{instance} of concept $C$.
   %This relation is used in the type system, in which $\instance$ assigns one or more concepts to every atom in the data set.
   %Since $\instance$ is a relation and every relation is a set of pairs,
   %set operators $\cup$, $\cap$, and $-$ can be used on $\instance$.

   Relations serve to organize and store data, allowing developers to represent facts.
   In this paper, variable $r$ represents relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
   Every relation $r$ has a name, a source concept, and a target concept.
   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
   The part $\pair{A}{B}$ is called the \define{signature} of the relation.

@@

   \begin{definition}[Data set]
   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance\subseteq\Pair{\Atoms}{\Concepts}$ that satisfies:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
\label{eqn:wellTypedEdge}
\end{eqnarray}
\end{definition}
   Looking at the example,
   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that the Person Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\popF{r}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, also known as the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop-rel}
\end{equation}
   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a data set $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
   so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
%   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
%\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\instance_{\dataset}A\ \wedge\ b\instance_{\dataset}B\]
%   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
%
   We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
   and expressions. We also define the difference of populations, in equation~\ref{eqn:pop-expr}, both for relations and concepts:
\begin{align}
   \pop{C}{\dataset}&= \{ x\mid\ x\ \instance_{\dataset}\ C\}
\label{eqn:pop-concept}\\
   \pop{x-y}{\dataset}&= \pop{x}{\dataset} - \pop{y}{\dataset}
\label{eqn:pop-expr}
\end{align}

@@

   In case $u$ is a transactional invariant,
   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$ such that $\pair{A}{B}=\sign{u}$.
   This requires that adding the pair to that relation solves the violation:
\begin{equation}
   \begin{array}[t]{l}
      (a,b)\in \viol{u}{\pair{\triples}{\instance}}\, \Longrightarrow \\
      \quad \viol{u}{\pair{\triples \cup \{\triple{a}{\declare{n}{A}{B}}{b}\}}{\instance}} = \viol{u}{\pair{\triples}{\instance}} - \{(a,b)\}
   \end{array}
   % SJC's remark:  \instance \cup \{\pair{a}{A},\pair{b}{B}\} = \instance ivm eqn:wellTypedViolation
\label{eqn:transaction}
\end{equation}
   It is obvious that not every conceivable constraint can satisfy this equation.
   So, we assume that the compiler restricts the set of transactional invariants to those that satisfy equation~\ref{eqn:transaction}.
   As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
   We call this the \define{enforced relation} of transactional invariant $u$:
\begin{equation}
   \enfRel{u}=\declare{n}{A}{B}
\end{equation}
   Let us denote a transactional invariant as $r\mapsfrom\violC{u}$ or equivalently $r \mapsfrom \lambda \dataset.~\viol{u}{\dataset}$,
   in which $r=\enfRel{u}$.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define information systems by enumerating the requirements.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}).
   We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
\item Triples in the data set must have their relation mentioned in the schema:
   \begin{eqnarray}
      \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
\item The system keeps any transactional invariant $u$ satisfied by adding violations to the relation $\enforceC{u}$ (\ref{eqn:transaction}).
\item All invariants must remain satisfied:
   \begin{align}
      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
\end{itemize}
\end{definition}
   We assume that a deployment will fail if these requirements are not satisfied.

@@

\section{Proof of Concept}
\label{sct:PoC}
   By way of proof of concept (PoC), we have built a migration system in Ampersand.
   To demonstrate it in the context of this paper, the existing system, $\pair{\dataset}{\schema}$, is rather trivial.
   It has no constraints and just one relation, $\declare{r}{A}{B}$.
   Its population is $A=\{a_1,a_2,a_3\}$, $B=\{b_1\}$, and $\pop{r}{\dataset}=\{\pair{a_1}{b_1}\}$.
   The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
   The schema of the migration system, $\schema_\migrsys$, follows from definition~\ref{eqn:schema migrsys}.

@@

   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial data set, $\dataset_{\infsys'}$.
   We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
   Together, the schema and the intial data set forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
   With these assumptions in place,
   we must verify that:
\begin{enumerate}
\item After predeployment, the migration system $\migrsys$ copies the designated triples to the desired system in finite time.
\item At MoT, $\migrsys$ satisfies the definition of information system (\ref{def:information system}), especially that it has no violations of blocking invariants (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
\item Once business actors have fixed the new business invariants, i.e. when the condition for the MoC (requirement~\ref{eqn:readyForMoC}) is satisfied,
   the behavior of $\migrsys$ and the desired system is identical, so moving the ingress from $\migrsys$ to the desired system is not noticable for end-users.
\item The old system has become redundant, so we can remove $\infsys$ and $\migrsys$.
\end{enumerate}
   Let us discuss these points one by one.

@@

\section{Conclusion}
\label{sct:Conclusions}
   We have shown that it is possible to generate a migration system from the schemas of an existing system and a desired system.
   We have defined a migration system such that:
   \begin{itemize}
   \item after deploying the migration system, data from the existing system will be copied automatically,
   \item the migration system behaves as a relaxed version of the desired system,
   \item the migration engineer can introduce transactional invariants in the migration system to address data pollution,
   \item there is no down time in switching from the existing system to the migration system at the MoT,
   \item the number of violations the business needs to fix is finite and known at the MoT,
   \item there is no down time in switching from the migration system to the desired system at the MoC,
   \item the migration system is equivalent to the desired system at the MoC.
   \end{itemize}

   The data migration we propose is certainly not the only approach one could think of.
   However, we have not come across other approaches that allow changing the schema in the presence of constraints.
   As such, we cannot compare our approach against other approaches yet.
   We envision that one day there will be multiple approaches for migration under a changing schema to choose from.
   For now, our next step is to implement the approach shown here into Ampersand.

   This work does not consider what to do about (user) interfaces yet.
   Instead, it models events by assuming that any change to the data set can be achieved.
   In practice, such changes need to be achieved through interfaces.
   Most Ampersand systems indeed allow the users of the system to edit the data set quite freely through the interfaces.
   This has allowed us to set aside interfaces in this paper for now.
   Nevertheless, we expect that an analysis of interfaces may yield insights about event streams,
   which are further building blocks to a foundational theory of information systems.
   But that is future work.

>>>>

\section{Introduction}
\label{sct:Introduction}
   In practice, information systems%
   \footnote{In the sequel, the word ``system'' refers to the phrase ``information system'', to simplify the language a little.}
   may live for many years.
   After they are built, they need to be updated regularly to keep up with changing requirements in a dynamically evolving environment.
   Schema changes cannot always be avoided when updating software, so a \define{schema-changing data migration} (SCDM) will be necessary from time to time.
   For example, adding or removing a column to a table in a relational database adds to the complexity of migrating data.
   Even worse, if a system invariant changes, some of the existing data in the system may violate the new invariant.
   In practice, data migrations typically follow Extract-Transform-Load (ETL) patterns~\cite{Theodorou2017},
   for which many tools are available.
   However, ETL tools typically provide little support for invariants that change, forcing development teams to write code.
   In a world where automation of the software process is resulting in higher productivity, more frequent releases, and better quality,
   SCDMs should not stay behind.
   Roughly half of the DevOps~\cite{BassWeberZhu15} teams that responded in a worldwide survey in 2023~\cite{HumanitecDevOps2023} are deploying software more frequently than once per day.
   Obviously, these deployments are mostly updates of existing systems.
   The risk and effort of SCDMs explains why these teams try to avoid schema changes in the first place.
   We believe that software teams that realize high update frequencies need better tools to perform SCDMs reliably with zero-downtime.
   We want them to design, implement, and deploy SCDMs in the flow of their DevOps practice.

   Data migration for other purposes than schema change has been described in the literature.
   For instance, if a data migration is done for switching to another platform or to different technology,
   e.g.~\cite{Gholami2016,Bisbal1999},
   migration engineers can and will avoid schema changes and functionality changes
   to avoid introducing new errors in an otherwise error-prone migration process.
   In another example, Ataei, Khan, and Walkingshaw~\cite{Ataei2021,Walkingshaw2014} define a migration as a variation between two data structures.
   They show how to unify databases with slight variations by preserving all variations in one comprehensive database.
   This does not cater for schema changes, however.
   Then there are SCDMs in situations without a schema or with an implicit schema, e.g.~\cite{Hillenbrand2022}.
   Such situations lack the error-preventing power that explicit schemas bring during the development of software.
   All errors a schema can prevent at compile time must then be compensated by runtime checks,
   which increases the likelihood of end-users getting error messages.
   This requires versioned storage of production data and an overhead in performance.
   That is why this paper focuses on SCDMs for systems with an explicit schema.
   The prototypical use case for that is to release updates of information systems in production,
   where the semantic integrity of data must be preserved across schema changes.
   Another use case is application integration for multiple dispersed data sources with explicit schemas.

@@

   The next section analyzes SCDMs with an eye on zero downtime and data quality.
   It sketches the outline of a procedure for SCDMs.
   Section~\ref{sct:Definitions} formalizes the concepts that we need to define the procedure.
   Section~\ref{sct:Generating} defines the migration system that automates SCDMs.
   Section~\ref{sct:PoC} exhibits the prototype of a migration system, which we used to verify our approach experimentally.
   We have used the language Ampersand for this purpose,
   because its syntax and semantics correspond directly to the definitions in section~\ref{sct:Definitions}.
   Finally, section~\ref{sct:Validation} discusses the validation of our approach by showing that all requirements are met.

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
   The current section yields a procedure for migrating data from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and at different moments, collectively form what we can call ``the business''.
   The data in the system constitutes the collective memory of the business.

   Constraints represent the elementary truths of a business,
   for example: ``{\em To submit a comment to the website of ACME Inc., the user must be registered.}''
   By formally defining constraints, designers can restrict a data set.
   We call a constraint {\em invariant} when it is implemented in an information system
   to ensure that the data set satisfies this constraint at all times as long as it is deployed.
   Thus, constraints implement the business semantics of the data.

   There are many ways to make sure that the data set satisfies a constraint.
   In this paper, we distinguish just three different kinds of enforcement:
\begin{enumerate}
\item Blocking invariant\\
   A \define{blocking invariant} is a constraint that is always satisfied in a system.
   It serves to constrain the data set at runtime.
   When the data set changes in a way that violates a blocking invariant, the system produces an error message and refuses the change.
\item Transactional invariant\\
   The classical database transaction can be understood as a \define{transactional invariant},
   which the system keeps satisfied by adding or deleting triples to the data set.
   As soon as the data violates this constraint, the system restores it without human intervention.
   So, the outside world experiences this as a constraint that is always satisfied, i.e.~an invariant.
\item Business constraint\\
   A \define{business constraint} is a constraint that users can violate temporarily until someone restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").
   That takes some time, during which the constraint is violated.
   So, we do not consider business constraints to be invariants.
\end{enumerate}
   Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
   Of the three types of constraint, only two are invariants.

   \begin{figure}[bht]
      \begin{center}
        \includegraphics[scale=0.8]{figures/existing_system.pdf}
      \end{center}
   \caption{Anatomy of an information system}
   \label{fig:pre-migration}
   \end{figure}

   Figure~\ref{fig:pre-migration} depicts the situation before migration.
   An existing application service ingests traffic through an ingress and persists data in a data set, which is typically a database.
   Our research assumes that the structure and business semantics are represented in a schema, from which the system is generated.
   Actors (both users and computers) are changing the data in a system continually.
   The state of the system is represented by a data set, typically represented in some form of persistent store such as a database.
   Events that the system detects may cause the state to change.
   To keep our approach technology-independent, we assume that data sets contain triples.
   This makes our approach valid for every kind of database that can be represented by triples,
   including SQL databases, object-oriented databases, graph databases, triple stores, and other no-SQL databases.
\subsection{Ampersand}
   We employ Ampersand as a prototyping language to demonstrate our approach.
   Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our approach.
   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
   The systems they generate keep invariants satisfied and alert users to violations of business constraints.
   The absence of imperative code in Ampersand scripts enhances reasoning about the system,
   while its static typing~\cite{vdWoude2011} yields the established benefits in software engineering processes~\cite{HanenbergKRTS14,Petersen2014}.
   Constraints carry business semantics, which makes ``preserving the meaning as much as possible'' explicit.
   An Ampersand script provides just enough information to generate a complete system,
   allowing extraction of a classical database schema (i.e., data structure plus constraints) from the Ampersand script.

@@

   Atoms serve as data elements.
   All atoms are taken from an infinite set called $\Atoms$.
   %
   Concepts are names that group atoms of the same type.
   All concepts are taken from an infinite set $\Concepts$.
   %$\Concepts$ and $\Atoms$ are disjoint.
   For example, a developer might choose to classify \atom{Peter} and \atom{Melissa} as \concept{Person},
   and \atom{074238991} as a \concept{TelephoneNumber}.
   In this example, \atom{Peter}, \atom{Melissa} and \atom{074238991} are atoms and
   \concept{Person} and \concept{TelephoneNumber} are concepts.
   In the sequel, variables $A$, $B$, and $C$ will represent concepts, and variables $a$, and $b$ represent \emph{atoms}.
   %
   The relation $\instance:\Pair{\Atoms}{\Concepts}$ relates atoms to concepts.
   The term $a\instance C$ means that atom $a$ is an \emph{instance} of concept $C$.
   %This relation is used in the type system, in which $\instance$ assigns one or more concepts to every atom in the data set.
   %Since $\instance$ is a relation and every relation is a set of pairs,
   %set operators $\cup$, $\cap$, and $-$ can be used on $\instance$.

   Relations serve to organize and store data, allowing developers to represent facts.
   In this paper, variables $r$ and $s$ represent relations\footnote{Some readers might like to read `relation symbol' where we write `relation'.}.
   All relations are taken from an infinite set $\Rels$.
   $\Rels$ is disjoint from $\Concepts$ and $\Atoms$.
   Every relation $r$ has a name, a source concept, and a target concept.
   The notation $r=\declare{n}{A}{B}$ denotes that relation $r$ has name $n$, source concept $A$, and target concept $B$.
   The part $\pair{A}{B}$ is called the \define{signature} of the relation.

@@

   \begin{definition}[Data set]
   A data set $\dataset$ is a tuple $\pair{\triples}{\instance}$ with finite $\triples \subseteq {\Triple{\Atoms}{\Rels}{\Atoms}}$ and $\instance\subseteq\Pair{\Atoms}{\Concepts}$ that satisfies:
\begin{eqnarray}
   \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&a\instance A\ \wedge\ b\instance B
\label{eqn:wellTypedEdge}
\end{eqnarray}
\end{definition}
   All datasets are taken from an infinite set $\Dataset$.
   Looking at the example,
   equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that the Person Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   The notations $\triples_{\dataset}$ and $\instance_{\dataset}$ are used to disambiguate $\triples$ and $\instance$ when necessary.
   To save writing in the sequel, the notation $a\ r\ b$ means that $\triple{a}{r}{b}\in\triples$.

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\popF{r}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, also known as the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop-rel}
\end{equation}
   Note that the phrase ``pair $\pair{a}{b}$ is in relation $r$'' means that there is a data set $\dataset$ in which $\pair{a}{b}\in\pop{r}{\dataset}$,
   so the phrase ``triple $\triple{a}{r}{b}$ is in $\dataset$'' means the same thing.
%   Equation~\ref{eqn:wellTypedEdge} implies that for every data set $\dataset$:
%\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ a\instance_{\dataset}A\ \wedge\ b\instance_{\dataset}B\]
%   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
%
   We overload the notation $\popF{}$ so we can use it on concepts $\popF{C}:\Dataset\rightarrow\powerset{\Atoms}$
   and expressions. We also define the difference of populations, in equation~\ref{eqn:pop-expr}, both for relations and concepts:
\begin{align}
   \pop{C}{\dataset}&= \{ x\mid\ x\ \instance_{\dataset}\ C\}
\label{eqn:pop-concept}\\
   \pop{r-s}{\dataset}&= \pop{r}{\dataset} - \pop{s}{\dataset}
\label{eqn:pop-expr}
\end{align}

@@

   In case $u$ is a transactional invariant,
   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$ such that $\pair{A}{B}=\sign{u}$.
   This requires that adding the pair to that relation solves the violation:
\begin{equation}
   \begin{array}[t]{l}
      \pair{a}{b}\in \viol{u}{\pair{\triples}{\instance}}\, \Longrightarrow \\
      \quad \viol{u}{\pair{\triples \cup \{\triple{a}{\declare{n}{A}{B}}{b}\}}{\instance}} = \viol{u}{\pair{\triples}{\instance}} - \{\pair{a}{b}\}
   \end{array}
   % SJC's remark:  \instance \cup \{\pair{a}{A},\pair{b}{B}\} = \instance ivm eqn:wellTypedViolation
\label{eqn:transaction}
\end{equation}
   It is obvious that not every conceivable constraint can satisfy this equation.
   So, we assume that the compiler restricts the set of transactional invariants to those that satisfy equation~\ref{eqn:transaction}.
   As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
   We call this the \define{enforced relation} of transactional invariant $u$:
\begin{equation}
   \enfRel{u}=\declare{n}{A}{B}
\end{equation}
   Let us denote a transactional invariant as $r\mapsfrom\violC{u}$ or equivalently $r \mapsfrom \lambda \dataset.~\viol{u}{\dataset}$,
   in which $r=\enfRel{u}$.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define information systems by enumerating the requirements.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}).
   We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} through~\ref{eqn:enforcement-has-type}).
\item Triples in the data set must have their relation mentioned in the schema:
   \begin{eqnarray}
      \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
\item The system keeps any transactional invariant $u$ satisfied by adding violations to the relation $\enforceC{u}$ (\ref{eqn:transaction}).
\item All invariants must remain satisfied:
   \begin{align}
      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
\end{itemize}
\end{definition}
   We assume that a deployment will fail if these requirements are not satisfied.

@@

\section{Proof of Concept}
\label{sct:PoC}
   By way of proof of concept (PoC), we have built a tiny migration system in Ampersand.
   It is tiny only to fit it in the page limit of this paper.
   The existing system, $\pair{\dataset}{\schema}$, has no constraints and just one relation, $\declare{r}{A}{B}$.
   Its population is $A=\{a_1,a_2,a_3\}$, $B=\{b_1\}$, and $\pop{r}{\dataset}=\{\pair{a_1}{b_1}\}$.
   The desired system contains one blocking invariant, which is the totality of $\declare{r}{A}{B}$.
   The schema of the migration system, $\schema_\migrsys$, follows from definition~\ref{eqn:schema migrsys}.

@@

   The initial situation consists of one existing system $\infsys$ and one desired system $\infsys'$ of which we have a schema, $\schema_{\infsys'}$ and an initial data set, $\dataset_{\infsys'}$.
   We may assume that $\infsys$ satisfies definition~\ref{def:information system} because it is a system in production.
   Also, $\schema_{\infsys'}$ satisfies equations~\ref{eqn:relationsIntroduceConcepts} through~\ref{eqn:enforcement-has-type} because the script of the desired system is type-correct.
   Together, the schema and the intial data set forms the desired system $\pair{\schema_{\infsys'}}{\dataset_{\infsys'}}$, which satisfies definition~\ref{def:information system}.
   With these assumptions in place,
   we must verify that:
\begin{enumerate}
\item After predeployment, the migration system $\migrsys$ copies the designated triples to the desired system in finite time.
\item At MoT, $\migrsys$ satisfies the definition of information system (\ref{def:information system}), especially that it has no violations of blocking invariants (requirement~\ref{eqn:satisfaction}) to ensure a successful deployment.
\item Once business actors have fixed the new business invariants, i.e. when the condition for the MoC (requirement~\ref{eqn:readyForMoC}) is satisfied,
   the behavior of $\migrsys$ and the desired system is identical, so moving the ingress from $\migrsys$ to the desired system is not noticable for end-users.
\item The old system has become redundant, so we can remove $\infsys$ and $\migrsys$.
\end{enumerate}
   Let us discuss these points one by one.

@@

\section{Conclusion}
\label{sct:Conclusions}
   We have shown that it is possible in theory to generate a migration system from the schemas of an existing system and a desired system.
   We have defined a migration system such that:
   \begin{itemize}
   \item after deploying the migration system, data from the existing system will be copied automatically,
   \item the migration system behaves as a relaxed version of the desired system,
   \item the migration engineer can introduce transactional invariants in the migration system to address data pollution,
   \item there is no down time in switching from the existing system to the migration system at the MoT,
   \item the number of violations the business needs to fix is finite and known at the MoT,
   \item there is no down time in switching from the migration system to the desired system at the MoC,
   \item the migration system is equivalent to the desired system at the MoC.
   \end{itemize}

   We intend to address the following issues in future work:
   \begin{itemize}
   \item Building the migration system generator in Ampersand, so it can be used in practice.
   \item Investigating the possibility of in-situ migration with changing schemas, to enable incremental development in production.
   \item Extending the theory to incorporate interfaces, so we can reason about the behavior of the system as a whole.
   \item Keep looking for other approaches to migration under a changing schema, to compare them with the approach shown here.
   So far, we have not found any, so we cannot compare our approach against other approaches yet.
   \end{itemize}
